---
title: "Preprocess data"
author: "Adam J Campbell"
date: "14/07/2020"
output: html_document
---

This file will help process the census data into the scoring information in the Shiny app (app.R).

You will need to download the individual shapefiles and census data.


Load your packages
```{r}
library(sf)
library(ggplot2)
library(dplyr)
library(rmapshaper)
library(leaflet)
library(tidyr)
```


load commuting data and pre-process 
```{r}
commute_data <- read.csv("raw_data/statsnz2018-census-main-means-of-travel-to-work-by-statistical-area-CSV/2018-census-main-means-of-travel-to-work-by-statistical-area.csv", na.strings = "-999") %>%  # load data treat -999 as NA
  rename(ID_RES = ï..SA2_code_usual_residence_address ) %>%  # do some basic renamings
  rename(ID_WORK = SA2_code_workplace_address) %>%
  select(-c("SA2_name_usual_residence_address",   # drop columns you don't need
    "SA2_usual_residence_easting",
    "SA2_usual_residence_northing",
    "SA2_name_workplace_address",
    "SA2_workplace_easting",
    "SA2_workplace_northing")) %>%
  mutate(ID_RES = as.character(ID_RES)) %>%  # change columns to character type for joins
  mutate(ID_WORK = as.character(ID_WORK)) 
```


Read NonDCC shapefile
```{r}
SA2_nonDCC <- st_read('raw_data/SA2_NON_DCC.shp') %>% # read data
  mutate(ID = "000100") %>%
  mutate(ID_NAME = "Other New Zealand") %>%
  rename(LAND_AREA = LAND_AREA_) %>%
  rename(TOT_AREA = AREA_SQ_KM) %>%
  rename(Shape_Leng = SHAPE_Leng) %>%
  select(-c("TA2020_V1_", "TA2020_V_1", "TA2020_V_2")) %>%
  ms_simplify(keep = 0.01) 

SA2_nonDCC <- SA2_nonDCC[c(4,5,1,2,3)]
```

Read Statistic area shapefile
```{r}
SA2 <- st_read('data/DCC_SA2.shp') %>% # read data
  #ms_simplify(keep = 0.2) %>% # simplify 
  #st_transform("+proj=longlat +datum=WGS84") %>% # change projection to lat/long
  rename(ID = SA22018_V1) %>%
  rename(ID_NAME = SA22018__1) %>%
  rename(LAND_AREA = LAND_AREA_) %>%
  rename(TOT_AREA = AREA_SQ_KM) %>%
  ms_simplify(keep = 0.2)
  

SA2_data <- st_drop_geometry(SA2)  # make a df of just data
```

This created two data frames `work_filter` and `res_filter` capturing people who live XOR work in Dunedin.

```{r}
Test_id <- 343600
NON_DCC_ID <- "000100"
commute_data_na <- commute_data
commute_data_na[is.na(commute_data_na)] <- 0


# people who live in Dunedin but commute elsewhere
work_filter <- commute_data_na%>%
  filter(ID_RES %in% SA2_data$ID) %>%
  group_by(ID_RES) %>%
  #filter(ID_RES == Test_id) %>%
  filter(!ID_WORK %in% SA2_data$ID) %>%
  summarise(
    ID_RES = ID_RES,
    ID_WORK= NON_DCC_ID,
    Work_at_home = sum(Work_at_home),
    Drive_a_private_car_truck_or_van = sum(Drive_a_private_car_truck_or_van),
    Drive_a_company_car_truck_or_van = sum(Drive_a_company_car_truck_or_van),
    Passenger_in_a_car_truck_van_or_company_bus = sum(Passenger_in_a_car_truck_van_or_company_bus),
    Public_bus = sum(Public_bus),
    Train = sum(Train),
    Bicycle = sum(Bicycle),
    Walk_or_jog = sum(Walk_or_jog),
    Ferry = sum(Ferry),
    Other = sum(Other),
    Total = sum(Total)) %>%
  distinct()

# people who live elsewhere but work in Dunedin
res_filter <- commute_data_na%>%
  filter(ID_WORK %in% SA2_data$ID) %>%
  group_by(ID_WORK) %>%
  #filter(ID_RES == Test_id) %>%
  filter(!ID_RES %in% SA2_data$ID) %>%
  summarise(
    ID_RES = NON_DCC_ID,
    ID_WORK= ID_WORK,
    Work_at_home = sum(Work_at_home),
    Drive_a_private_car_truck_or_van = sum(Drive_a_private_car_truck_or_van),
    Drive_a_company_car_truck_or_van = sum(Drive_a_company_car_truck_or_van),
    Passenger_in_a_car_truck_van_or_company_bus = sum(Passenger_in_a_car_truck_van_or_company_bus),
    Public_bus = sum(Public_bus),
    Train = sum(Train),
    Bicycle = sum(Bicycle),
    Walk_or_jog = sum(Walk_or_jog),
    Ferry = sum(Ferry),
    Other = sum(Other),
    Total = sum(Total)) %>%
  distinct()

# rearrange column names
res_filter <- res_filter[c(2,1,3,4,5,6,7,8,9, 10, 11, 12, 13)]
```

Get people who live AND work in Dunedin
```{r}
commute_live_and_work <- commute_data_na %>%
  filter(ID_RES %in% SA2_data$ID & ID_WORK %in% SA2_data$ID)

commute_clip <- rbind(commute_live_and_work, work_filter) %>%
  rbind(res_filter)
```


Get truncated SA2
```{r}
residence_join <- left_join(commute_data, SA2_data, by = c("ID_RES"= "ID"))
workplace_join <- left_join(commute_data, SA2_data, by = c("ID_RES"= "ID"))

unique_residence <- unique(residence_join$ID_RES)
unique_workplace <- unique(workplace_join$ID_WORK)

new <- unique_workplace[!unique_workplace %in% unique_residence]
new2 <- unique_residence[!unique_residence %in% unique_workplace]

new3 <- SA2_data[!SA2_data$ID %in% commute_data$ID_WORK, ]

new4 <- SA2_data[SA2_data$ID %in% commute_data$ID_WORK, ]

SA2_truncated <- SA2[SA2$ID %in% new4$ID, ]

SA2_combined <- rbind(SA2_truncated, SA2_nonDCC)
```


```{r}
ggplot(data=SA2_combined) +
  geom_sf()
```

SA2_combined has the data I want
```{r}
SA2_combined <- SA2_combined %>%
  #ms_simplify(keep = 0.2) %>% # simplify 
  st_transform("+proj=longlat +datum=WGS84")  # change projection to lat/long
```

Coming/going analysis
```{r}
workers_coming<-commute_clip %>%
  group_by(ID_WORK) %>%
  summarise(coming = sum(Total) - sum(Work_at_home), staying = sum(Work_at_home), total_workers = sum(Total))
workers_going <-commute_clip %>%
  group_by(ID_RES) %>%
  summarise(going = sum(Total) - sum(Work_at_home))

coming_going <- full_join(workers_coming, workers_going, by = c("ID_WORK" = "ID_RES"))%>%
  mutate(net = coming - going)

```

drop unneeded columns and rename columns
```{r}
commute_clip <-
  commute_clip %>% 
  rename("Work at Home" = Work_at_home) %>%
  rename("On foot" = Walk_or_jog) %>%
  rename("Bus" = Public_bus) %>%
  rename("Passenger" = Passenger_in_a_car_truck_van_or_company_bus) %>%
  mutate("Drive" = Drive_a_private_car_truck_or_van + 
      Drive_a_company_car_truck_or_van) %>%
  select(-c("Train", 
    "Ferry", 
    "Drive_a_private_car_truck_or_van", 
    "Drive_a_company_car_truck_or_van"))
```

compute sums and scores
```{r}
commuting_from_total_sums <- commute_clip %>%
  mutate(Drive_all = Drive + Passenger ) %>%
  summarise(
    drive_pct_total = sum(Drive_all)/ sum(Total),
    bus_pct_total = sum(Bus)/ sum(Total),
    bike_pct_total = sum(Bicycle)/ sum(Total),
    on_foot_pct_total = sum(`On foot`) / sum(Total),
    wfh_pct_total = sum(`Work at Home`) /sum(Total)
    )


commuting_from_pcts <- commute_clip %>% 
  mutate(Drive_all = Drive + Passenger ) %>%
  group_by(ID_RES) %>%
  summarise(
    total_sum = sum(Total),
    drive_sum = sum(Drive_all),
    drive_pct = sum(Drive_all)/sum(Total),
    bus_sum = sum(Bus),
    bus_pct = sum(Bus)/sum(Total),
    bike_sum = sum(Bicycle),
    bike_pct = sum(Bicycle)/sum(Total),
    onfoot_sum = sum(`On foot`),
    onfoot_pct = sum(`On foot`)/sum(Total),
    wfh_sum = sum(`Work at Home`),
    wfh_pct = sum(`Work at Home`)/sum(Total),
    ) %>%
  mutate(
    drive_ifaverage = commuting_from_total_sums$drive_pct_total * total_sum,
    bus_ifaverage = commuting_from_total_sums$bus_pct_total * total_sum,
    bike_ifaverage = commuting_from_total_sums$bike_pct_total * total_sum,
    onfoot_ifaverage = commuting_from_total_sums$on_foot_pct_total * total_sum,
    wfh_ifaverage = commuting_from_total_sums$wfh_pct_total * total_sum
    ) %>%
  mutate(
    drive_diff = drive_sum - drive_ifaverage,
    bus_diff = bus_sum - bus_ifaverage,
    bike_diff = bike_sum - bike_ifaverage,
    onfoot_diff = onfoot_sum - onfoot_ifaverage,
    wfh_diff = wfh_sum - wfh_ifaverage,
    direction = "from"
    ) %>%
  rename(ID = ID_RES)

commuting_to_pcts <- commute_clip %>% 
  mutate(Drive_all = Drive + Passenger ) %>%
  group_by(ID_WORK) %>%
  summarise(
    total_sum = sum(Total),
    drive_sum = sum(Drive_all),
    drive_pct = sum(Drive_all)/sum(Total),
    bus_sum = sum(Bus),
    bus_pct = sum(Bus)/sum(Total),
    bike_sum = sum(Bicycle),
    bike_pct = sum(Bicycle)/sum(Total),
    onfoot_sum = sum(`On foot`),
    onfoot_pct = sum(`On foot`)/sum(Total),
    wfh_sum = sum(`Work at Home`),
    wfh_pct = sum(`Work at Home`)/sum(Total),
    ) %>%
  mutate(
    drive_ifaverage = commuting_from_total_sums$drive_pct_total * total_sum,
    bus_ifaverage = commuting_from_total_sums$bus_pct_total * total_sum,
    bike_ifaverage = commuting_from_total_sums$bike_pct_total * total_sum,
    onfoot_ifaverage = commuting_from_total_sums$on_foot_pct_total * total_sum,
    wfh_ifaverage = commuting_from_total_sums$wfh_pct_total * total_sum
    ) %>%
  mutate(
    drive_diff = drive_sum - drive_ifaverage,
    bus_diff = bus_sum - bus_ifaverage,
    bike_diff = bike_sum - bike_ifaverage,
    onfoot_diff = onfoot_sum - onfoot_ifaverage,
    wfh_diff = wfh_sum - wfh_ifaverage,
    direction = "to"
    ) %>%
  rename(ID = ID_WORK)

commuting_pcts <-rbind(commuting_from_pcts, commuting_to_pcts)


pivoted_commuting <-commuting_pcts %>%
  pivot_longer(
    cols = -c(direction,ID, total_sum),
    names_to = c("mode", "type"),
    names_sep = "_",
    values_to = "value"
  )


```


save data
```{r}
st_write(SA2_combined, 'data/SA2_combined.shp', append =FALSE)
saveRDS(commute_clip, file="data/commute_clip")
saveRDS(coming_going, file="data/coming_going")
saveRDS(commuting_pcts, file = "data/commute_pct")
saveRDS(pivoted_commuting, file = "data/commute_pivot")
```
